---
title: "Uneven Green: Urban Greenspace Inequality in South Asia"
author: "Sonika Mahat"
institute: "American University, SIS-750: Data Analysis"
date: "2025-04-09"
output: 
  beamer_presentation:
    theme: "Berlin"
    colortheme: "whale"
  fontsize: 9pt
  classoption: "aspectratio=169"
---
```{r setup, include=FALSE}
#Packages
library(knitr)
library(rmdformats)
library(tidyverse)
#Global options
knitr::opts_chunk$set(
  echo = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center')
```
#  Overview: Motivation and Research Question

##  Motivation

**Why is Greenspace Important in Urban Areas?**

-  Improves air quality, cools cities, and supports biodiversity.  
-  Promotes physical and mental health, social cohesion, and resilience

In South Asia, urban greenspace is under increasing pressure from rapid population growth and uneven planning.

This study explores how greenspace differs **across countries** and **between cities**, and whether those differences relate to **HDI levels**

-  Note: HDI = Human Development Index

##  Research Question and Key Findings

**Research Question:**  

-  How does urban greenspace vary across South Asian countries?
-  What patterns emerge when we look at cities, not just country averages?
-  Are cities in higher-HDI countries more likely to have greener urban areas?

# Data

## Global Greenspace Indicator Dataset  

**Data Source:**  
-  Stowell, Jennifer. 2023. “Global Greenspace Indicator Dataset.” Harvard Dataverse. https://doi.org/doi:10.7910/DVN/TMWYHB.

**Scope for this Project:**  
-  Focused on cities in South Asia: India, Nepal, Bangladesh, Pakistan, Sri Lanka, Bhutan, Maldives, Afghanistan.

**Key variables used:**  
-  `annual_avg_2021`: NDVI score for urban greenspace in 2021  
-  `City`, `Country`: Used for both national and local comparisons  
-  `HDI_level`: Categorical variable for Human Development Index (Low/Medium/High)

# Analysis of Dataset: Analysis 1, 2, and 3

## Analysis 1: Greenspace by Country

**Research Question:** How does urban greenspace vary across South Asian countries?

**Method:** calculated the average 2021 NDVI score for cities within each country. This shows overall national patterns, but may mask city-level difference.

## Analysis 1: Greenspace by Country (cont.)

**Visualization:** Bar chart of mean NDVI (2021) for each South Asian country
```{r analysis1, fig.width=7, fig.height=3, out.width="0.8\\linewidth"}
# Load packages
library(tidyverse)

# Read the CSV file
greenspace_data <- read.csv("greenspace_data_share.csv", encoding = "latin1")

# Define South Asian countries
south_asia <- c("Afghanistan", "Bangladesh", "Bhutan", "India", 
                "Maldives", "Nepal", "Pakistan", "Sri Lanka")

# Filter to South Asian countries only
greenspace_sa <- greenspace_data %>%
  filter(Country %in% south_asia)

# Remove rows with missing 2021 NDVI values
greenspace_sa <- greenspace_sa %>%
  filter(!is.na(annual_avg_2021))

# Calculate average NDVI for each country (2021)
country_avg <- greenspace_sa %>%
  group_by(Country) %>%
  summarise(mean_ndvi_2021 = mean(annual_avg_2021, na.rm = TRUE)) %>%
  arrange(desc(mean_ndvi_2021))

# Bar chart using average NDVI in 2021 by country
ggplot(country_avg, aes(x = reorder(Country, -mean_ndvi_2021), 
                        y = mean_ndvi_2021, 
                        fill = mean_ndvi_2021)) + 
  geom_bar(stat = "identity", width = 0.6, show.legend = FALSE) +
  scale_fill_gradient(low = "#c7e9c0", high = "#006d2c") +
  expand_limits(y = 0) +
  labs(
    title = "Average Urban Greenspace by Country (2021)",
    x = "Country",
    y = "Mean NDVI (2021)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(t = 10))
  )
```
**Takeaway:** Sri Lanka most green, Afghanistan and Maldives least.

## Analysis 2: Greenspace by City

**Research Question:** What patterns emerge when we look at individual cities, not just country averages?

**Method:** Identified the Top 20 cities in South Asia with the highest NDVI scores (2021). Each city is labeled with its country for comparison.

## Analysis 2: Greenspace by City (cont.)
**Visualization**:
```{r analysis2, fig.width=7, fig.height=3, out.width="0.8\\linewidth"}
# Load packages
library(tidyverse)

greenspace_sa <- greenspace_sa %>%
  mutate(city_country = paste0(City, " (", Country, ")"))

#top 20 cities of South Asia
top_20 <- greenspace_sa %>%
  arrange(desc(annual_avg_2021)) %>%
  slice_head(n = 20)

custom_colors <- c("India" = "#006d2c",       # dark green
                   "Sri Lanka" = "#a1d99b",
                   "Bangladesh" = "#c7e9c0") 

# Plot: Top 20 Cities with two-shade green scheme
ggplot(top_20, aes(x = reorder(city_country, annual_avg_2021), 
                   y = annual_avg_2021, fill = Country)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Top 20 Cities with Highest Urban Greenspace in South Asia (2021)",
    x = "City (Country)",
    y = "NDVI Score (2021)"
  ) +
  scale_fill_manual(values = custom_colors) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
**Takeaway:** The greenest cities are mostly in India, especially in the south. This shows the importance of looking beyond country averages to understand local variation in urban greenness

## Analysis 3: Greenspace by HDI Level
**Research Question:** Are cities in more developed (higher-HDI) South Asian countries more likely to have greener urban areas?  

**Method:** Grouped cities into Low, Medium, and High HDI categories. Compared urban greenspace levels (NDVI 2021) using a boxplot and tested for statistical significance with one-way ANOVA.

## Analysis 3: Greenspace by HDI Level (cont.)
**Visualization:**
```{r analysis3, fig.width=6, fig.height=3.5, out.width="0.8\\linewidth"}
# Run ANOVA test
anova_result <- aov(annual_avg_2021 ~ HDI_level, data = greenspace_sa)
anova_summary <- summary(anova_result)

# Extract p-value
p_value <- anova_summary[[1]]["Pr(>F)"][1, 1]
p_text <- ifelse(p_value < 0.001, "p < 0.001", paste0("p = ", round(p_value, 3)))

# Plot with green fill palette
ggplot(greenspace_sa, aes(x = HDI_level, y = annual_avg_2021, fill = HDI_level)) +
  geom_boxplot(show.legend = FALSE, outlier.color = "gray") +
  scale_fill_manual(values = c(
    "Low" = "#c7e9c0",       # light green
    "Medium" = "#74c476",    # medium green
    "High" = "#238b45"       # dark green
  )) +
  labs(
    title = "Distribution of Urban Greenspace by HDI Level (2021)",
    x = "HDI Level",
    y = "NDVI Score"
  ) +
  annotate("text", x = 2, y = max(greenspace_sa$annual_avg_2021) + 0.02, 
           label = p_text, size = 4.5, fontface = "italic") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
**Takeaway:** Higher-HDI countries are greener than lower-HDI.

##  Conclusion  

-  Urban greenspace is unevenly distributed across South Asia  
-  While Sri Lanka leads in national average, India has many of the greenest cities  
-  Cities in higher-HDI countries tend to have more greenspace, and this difference is statistically significant (ANOVA, p < 0.001)  
- Results suggest that development level matters, both nationally and locally in shaping urban environmental quality

#  Thank you